---
title: "QB Pocket Pressure Analysis (PRSS) — Rolling Window Comparison"
output:
  html_document:
    df_print: paged
---

## 1. setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(arrow)
library(deldir)
library(nflfastR)
library(zoo)
library(ggrepel)
```

---

## 2. load data

```{r load-data, message=FALSE}

tracking_data <- read_parquet("../data/processed/bdb-tracking-all.parquet")

plays   <- read_csv("../data/bdb/plays.csv",            show_col_types = FALSE)
players <- read_csv("../data/bdb/players.csv",           show_col_types = FALSE)
pff     <- read_csv("../data/bdb/pffScoutingData.csv",   show_col_types = FALSE)
games <- read_csv("../data/bdb/games.csv", show_col_types = FALSE)

```

---

## 3. data cleaning

```{r clean-data}
tracking_data <- left_join(tracking_data, 
                           games %>% select(gameId, week), 
                           by = "gameId")

# When nflId is NA, row corresponds to ball. (numeric)
tracking <- tracking_data %>%
  filter(!is.na(nflId)) %>%
  left_join(
    players %>% select(nflId, officialPosition, displayName),
    by = "nflId"
  )

# create 'real pressure' signals from pff
# pff_hit / pff_hurry / pff_sack =1
pff_pressure <- pff %>%
  filter(pff_role == "Pass Rush") %>%
  group_by(gameId, playId) %>%
  summarise(
    actual_pressure = as.integer(any(pff_hit == 1 | pff_hurry == 1 | pff_sack == 1, na.rm = TRUE)),
    n_rushers       = n(),
    .groups = "drop"
  )

# Pocket participants: pass blockers + pass rushers (QB added separately)
pff_pocket_ids <- pff %>%
  filter(pff_role %in% c("Pass Block", "Pass Rush"), !is.na(nflId)) %>%
  distinct(gameId, playId, nflId) %>%
  group_by(gameId, playId) %>%
  summarise(pocket_ids = list(unique(nflId)), .groups = "drop")

```

---

## 4. **PRSS(Pocket Rate of Space Shrinkage)**

```{r define-function}
fallback_positions <- c(
  "QB", "C", "G", "T", "TE", "FB", "RB", "WR",
  "DE", "DT", "NT", "ILB", "OLB", "MLB", "LB",
  "CB", "DB", "FS", "SS"
)
back_cap_yards <- 3
side_cap_yards <- 3

poly_area <- function(x, y) {
  if (length(x) < 3 || length(y) < 3) return(NA_real_)
  x2 <- c(x, x[1])
  y2 <- c(y, y[1])
  abs(sum(x2[-1] * y2[-length(y2)] - x2[-length(x2)] * y2[-1])) / 2
}

clip_polygon <- function(x, y, inside_fn, intersect_fn) {
  n <- length(x)
  if (n < 3) return(list(x = numeric(0), y = numeric(0)))

  out_x <- numeric(0)
  out_y <- numeric(0)

  for (i in seq_len(n)) {
    j <- if (i == 1) n else i - 1
    x1 <- x[j]; y1 <- y[j]
    x2 <- x[i]; y2 <- y[i]
    in1 <- inside_fn(x1, y1)
    in2 <- inside_fn(x2, y2)

    if (in2) {
      if (!in1) {
        p <- intersect_fn(x1, y1, x2, y2)
        out_x <- c(out_x, p[1]); out_y <- c(out_y, p[2])
      }
      out_x <- c(out_x, x2); out_y <- c(out_y, y2)
    } else if (in1) {
      p <- intersect_fn(x1, y1, x2, y2)
      out_x <- c(out_x, p[1]); out_y <- c(out_y, p[2])
    }
  }

  list(x = out_x, y = out_y)
}

clip_xmin <- function(x, y, xmin) {
  clip_polygon(
    x, y,
    inside_fn = function(px, py) px >= xmin,
    intersect_fn = function(x1, y1, x2, y2) {
      dx <- x2 - x1
      t <- if (abs(dx) < 1e-12) 0 else (xmin - x1) / dx
      c(xmin, y1 + t * (y2 - y1))
    }
  )
}

clip_xmax <- function(x, y, xmax) {
  clip_polygon(
    x, y,
    inside_fn = function(px, py) px <= xmax,
    intersect_fn = function(x1, y1, x2, y2) {
      dx <- x2 - x1
      t <- if (abs(dx) < 1e-12) 0 else (xmax - x1) / dx
      c(xmax, y1 + t * (y2 - y1))
    }
  )
}

clip_ymin <- function(x, y, ymin) {
  clip_polygon(
    x, y,
    inside_fn = function(px, py) py >= ymin,
    intersect_fn = function(x1, y1, x2, y2) {
      dy <- y2 - y1
      t <- if (abs(dy) < 1e-12) 0 else (ymin - y1) / dy
      c(x1 + t * (x2 - x1), ymin)
    }
  )
}

clip_ymax <- function(x, y, ymax) {
  clip_polygon(
    x, y,
    inside_fn = function(px, py) py <= ymax,
    intersect_fn = function(x1, y1, x2, y2) {
      dy <- y2 - y1
      t <- if (abs(dy) < 1e-12) 0 else (ymax - y1) / dy
      c(x1 + t * (x2 - x1), ymax)
    }
  )
}

clip_qb_polygon <- function(x, y, qb_x, qb_y, play_dir, back_cap, side_cap) {
  out <- list(x = x, y = y)
  dir <- tolower(as.character(play_dir))

  if (is.finite(qb_x) && dir == "right") out <- clip_xmin(out$x, out$y, qb_x - back_cap)
  if (is.finite(qb_x) && dir == "left")  out <- clip_xmax(out$x, out$y, qb_x + back_cap)

  if (is.finite(qb_y)) {
    out <- clip_ymin(out$x, out$y, qb_y - side_cap)
    out <- clip_ymax(out$x, out$y, qb_y + side_cap)
  }
  out
}

# k: rolling window size
compute_qb_area <- function(play_data, k, pocket_ids = NULL) {

  # Find the snap frame and the pass/ sack / spike frame
  snap_vals    <- play_data$frameId[play_data$event == "ball_snap"]
  release_vals <- play_data$frameId[play_data$event %in%
                    c("pass_forward", "qb_sack", "qb_spike")]

  if (length(snap_vals) == 0 || length(release_vals) == 0) {
    return(tibble(qb = NA_character_, max_neg_change = NA_real_))
  }

  snap_frame    <- min(snap_vals,    na.rm = TRUE)
  release_frame <- min(release_vals, na.rm = TRUE)

  if (!is.finite(snap_frame) || !is.finite(release_frame)) {
    return(tibble(qb = NA_character_, max_neg_change = NA_real_))
  }
  if (release_frame < snap_frame) {
    return(tibble(qb = NA_character_, max_neg_change = NA_real_))
  }

  play_data <- play_data %>%
    filter(frameId >= snap_frame, frameId <= release_frame)

  # QB name
  qb_name <- play_data %>%
    filter(officialPosition == "QB") %>%
    distinct(displayName) %>%
    pull() %>%
    paste(collapse = "/")

  frames <- unique(play_data$frameId)

  # we need k+1
  if (length(frames) < (k + 1)) {
    return(tibble(qb = qb_name, max_neg_change = NA_real_))
  }

  # calculate QB's Voronoi area frame by frame
  qb_areas <- map_dfr(frames, function(f) {
    cleaned_ids <- unique(pocket_ids[!is.na(pocket_ids)])
    has_pff_ids <- length(cleaned_ids) > 0

    frame_data <- if (has_pff_ids) {
      play_data %>%
        filter(
          frameId == f,
          !is.na(x), !is.na(y),
          officialPosition == "QB" | nflId %in% cleaned_ids
        ) %>%
        distinct(x, y, .keep_all = TRUE) %>%
        mutate(ptNum = row_number())
    } else {
      play_data %>%
        filter(
          frameId == f,
          !is.na(x), !is.na(y),
          officialPosition %in% fallback_positions
        ) %>%
        distinct(x, y, .keep_all = TRUE) %>%
        mutate(ptNum = row_number())
    }

    if (nrow(frame_data) < 3) return(tibble(frameId = f, qb_area = NA_real_))

    tryCatch({
      qb_seed <- frame_data %>%
        filter(officialPosition == "QB") %>%
        slice(1)
      if (nrow(qb_seed) == 0) return(tibble(frameId = f, qb_area = NA_real_))

      qb_x_seed <- qb_seed$x[[1]]
      qb_y_seed <- qb_seed$y[[1]]
      play_dir <- tolower(as.character(qb_seed$playDirection[[1]]))

      vor <- deldir(frame_data$x, frame_data$y, rw = c(0, 120, 0, 53.3))
      tiles <- tile.list(vor)
      idx <- qb_seed$ptNum[[1]]
      if (!is.finite(idx) || idx < 1 || idx > length(tiles)) {
        return(tibble(frameId = f, qb_area = NA_real_))
      }

      tile <- tiles[[idx]]
      capped <- clip_qb_polygon(
        tile$x, tile$y,
        qb_x = qb_x_seed,
        qb_y = qb_y_seed,
        play_dir = play_dir,
        back_cap = back_cap_yards,
        side_cap = side_cap_yards
      )
      qb_area <- poly_area(capped$x, capped$y)
      tibble(frameId = f, qb_area = ifelse(is.finite(qb_area), qb_area, NA_real_))
    }, error = function(e) tibble(frameId = f, qb_area = NA_real_))
  }) %>%
    arrange(frameId) %>%
    mutate(
      rate_change = c(NA, diff(qb_area)),
      rolling_avg = suppressWarnings(
        zoo::rollmean(rate_change, k = k, align = "left", fill = NA)
      )
    )

  max_neg <- suppressWarnings(min(qb_areas$rolling_avg, na.rm = TRUE))
  if (!is.finite(max_neg)) max_neg <- NA_real_

  tibble(qb = qb_name, max_neg_change = max_neg)
}
```

---

## 5.  k = 3, 5, 7, 10 

### k=5
```{r}
play_list <- tracking %>%
  group_by(gameId, playId) %>%
  nest() %>%
  left_join(pff_pocket_ids, by = c("gameId", "playId"))

results <- vector("list", nrow(play_list))

for (i in seq_len(nrow(play_list))) {
  if (i %% 100 == 0) message("Progress: ", i, "/", nrow(play_list))
  results[[i]] <- compute_qb_area(
    play_list$data[[i]],
    k = 5,
    pocket_ids = play_list$pocket_ids[[i]]
  )
}

all_summary_k5 <- bind_rows(results) %>%
  bind_cols(play_list %>% select(gameId, playId))


# save
saveRDS(all_summary_k5, "../data/processed/all_summary_k5.rds")
```

```{r}
final_data <- all_summary_k5 %>%
  filter(!is.na(max_neg_change), !is.na(qb)) %>%
  left_join(
    plays %>% select(gameId, playId, possessionTeam, 
                     passResult, prePenaltyPlayResult),
    by = c("gameId", "playId")
  ) %>%
  left_join(pff_pressure, by = c("gameId", "playId"))
```

```{r}
# plot1: PRSS Distribution
ggplot(final_data, aes(x = max_neg_change)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of PRSS (k=5)",
    x = "PRSS (negative = pocket collapsing)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# PRSS vs playResult
ggplot(final_data, aes(x = max_neg_change, y = prePenaltyPlayResult)) +
  geom_point(alpha = 0.2, color = "lightcoral") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "PRSS vs Play Result (yards gained)",
    x = "PRSS",
    y = "Yards Gained"
  ) +
  theme_minimal()

# correlation
cor(final_data$max_neg_change, final_data$prePenaltyPlayResult, use = "complete.obs")
```

```{r}
# k=5 
k5 <- all_summary_k5 %>% mutate(window = as.factor(5))

# k=3, 7, 10
window_sizes <- c(3, 7, 10)
all_results <- vector("list", length(window_sizes))

for (w in seq_along(window_sizes)) {
  k <- window_sizes[w]
  message("Computing k = ", k, " ...")
  
  results_k <- vector("list", nrow(play_list))
  for (i in seq_len(nrow(play_list))) {
    if (i %% 500 == 0) message("  k=", k, " Progress: ", i, "/", nrow(play_list))
    results_k[[i]] <- compute_qb_area(
      play_list$data[[i]],
      k = k,
      pocket_ids = play_list$pocket_ids[[i]]
    )
  }
  
  all_results[[w]] <- play_list %>%
    select(gameId, playId) %>%
    bind_cols(bind_rows(results_k)) %>%
    mutate(window = as.factor(k))
}

# combine all windows
all_summaries <- bind_rows(all_results) %>%
  bind_rows(k5)

saveRDS(all_summaries, "../data/processed/all_summaries_all_windows.rds")
```

## 6. review the k 
```{r}
# load
all_summaries <- readRDS("../data/processed/all_summaries_all_windows.rds")

all_summaries_final <- all_summaries %>%
  filter(!is.na(max_neg_change), !is.na(qb)) %>%
  left_join(
    plays %>% select(gameId, playId, possessionTeam, 
                     passResult, prePenaltyPlayResult),
    by = c("gameId", "playId")
  )

all_summaries_final %>%
  group_by(window) %>%
  summarise(
    n = n(),
    na_count = sum(is.na(max_neg_change)),
    mean_prss = mean(max_neg_change, na.rm = TRUE),
    sd_prss = sd(max_neg_change, na.rm = TRUE)
  )
```

## PRSS visualization
```{r}
# plot1: PRSS Distribution (4 windows comparison)
ggplot(all_summaries_final, aes(x = max_neg_change, fill = window, color = window)) +
  geom_density(alpha = 0.25) +
  xlim(-20, 10) +  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "PRSS Distribution by Rolling Window Size",
    x = "PRSS",
    y = "Density",
    fill = "Window (k)",
    color = "Window (k)"
  ) +
  theme_minimal(base_size = 13)
```
As the rolling window size increases, the PRSS distribution shifts rightward and becomes wider.
k=3 (red): The distribution is very narrow and tall, centered just below 0. The small window picks up every small fluctuation in pocket area, but those fluctuations cancel each other out quickly, so the final PRSS values cluster tightly around 0.
k=5 (purple) and k=7 (green): Progressively wider and slightly more shifted to the right. These windows smooth out frame-to-frame noise while still capturing genuine pocket collapse events.
k=10 (teal): The widest and most right-shifted distribution, with mean PRSS actually turning positive (+0.46). The window is so large that it averages out real pressure events, making plays look like the pocket was stable or even expanding when it actually collapsed.

There is a tradeoff between sensitivity and smoothness. k=3 is too reactive to noise, k=10 over-smooths and loses real pressure signals. k=5 or k=7 appears to be the sweet spot — capturing genuine pocket collapse while filtering out frame-to-frame measurement noise.
The next step is to check which k value produces the strongest correlation with yards gained, which will help confirm the optimal window size.

```{r}

# plot2：the correlation between yards and windows
all_summaries_final %>%
  group_by(window) %>%
  summarise(
    correlation = cor(max_neg_change, prePenaltyPlayResult, use = "complete.obs"),
    .groups = "drop"
  )

```
All four correlations are extremely weak (0.018 - 0.028), and there's no meaningful difference between window sizes.
What this tells us:
The choice of k barely matters for predicting yards gained at the play level. All windows perform essentially the same.

```{r}

# only k=5，filter scramble
prss_by_result <- all_summaries_final %>%
  filter(window == '5', passResult %in% c("C", "I", "S", "IN")) %>%
  mutate(passResult = factor(passResult, 
                             levels = c("S", "IN", "I", "C"),
                             labels = c("Sack", "Interception", "Incomplete", "Complete")))

# boxplot
ggplot(prss_by_result, aes(x = passResult, y = max_neg_change, fill = passResult)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "PRSS Distribution by Pass Result (k=5)",
    x = "Pass Result",
    y = "PRSS (more negative = more pressure)",
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# mean
prss_by_result %>%
  group_by(passResult) %>%
  summarise(
    mean_prss = mean(max_neg_change, na.rm = TRUE),
    median_prss = median(max_neg_change, na.rm = TRUE),
    n = n()
  )

```
## 6. join external data

```{r merge-data, message=FALSE}
library(nflfastR)

pbp <- load_pbp(2021) %>%
  mutate(
    gameId = as.numeric(old_game_id),
    playId = as.numeric(play_id)
  ) %>%
  select(gameId, playId, epa, yards_gained, cpoe, air_yards)


all_summaries_final <- all_summaries_final %>%
  left_join(pbp, by = c("gameId", "playId"))

sum(!is.na(all_summaries_final$epa))
```

## 7. prss vs epa
```{r}

all_summaries_final %>%
  group_by(window) %>%
  summarise(
    correlation = cor(max_neg_change, epa, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )
```
At the individual play level, PRSS shows very weak correlation with EPA across all rolling window sizes (r = 0.04–0.05), with no meaningful difference between k = 3, 5, 7, and 10.

This is not necessarily an indicator that PRSS is ineffective. Pocket geometry is only one of many factors that determine the outcome of a play — receiver route quality, defensive coverage, quarterback decision-making, and receiver performance all contribute independently to EPA. A quarterback may face severe pocket collapse yet still deliver a successful pass, or face minimal pressure yet throw an incompletion.

The consistency of correlations across all four window sizes also suggests that the choice of k does not meaningfully affect PRSS at the play level, implying the metric is relatively robust to this parameter.

The more appropriate level of analysis may be at the QB or team level, where averaging across many plays reduces the noise from individual play variance and allows the signal from pocket pressure to emerge more clearly.
---
