---
title: "QB Pocket Pressure Analysis (PRSS) — Rolling Window Comparison"
output: html_notebook
---

## 1. 加载库

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(arrow)
library(deldir)
library(nflfastR)
library(zoo)
library(ggrepel)
```

---

## 2. load data

```{r load-data, message=FALSE}

tracking_data <- read_parquet("../data/processed/bdb-tracking-all.parquet")

plays   <- read_csv("../data/bdb/plays.csv",            show_col_types = FALSE)
players <- read_csv("../data/bdb/players.csv",           show_col_types = FALSE)
pff     <- read_csv("../data/bdb/pffScoutingData.csv",   show_col_types = FALSE)
games <- read_csv("../data/bdb/games.csv", show_col_types = FALSE)

```

---

## 3. data cleaning

```{r clean-data}
tracking_data <- left_join(tracking_data, 
                           games %>% select(gameId, week), 
                           by = "gameId")

# When nflId is NA, row corresponds to ball. (numeric)
tracking <- tracking_data %>%
  filter(!is.na(nflId)) %>%
  left_join(
    players %>% select(nflId, officialPosition, displayName),
    by = "nflId"
  )

# create 'real pressure' signals from pff
# pff_hit / pff_hurry / pff_sack =1
pff_pressure <- pff %>%
  filter(pff_role == "Pass Rush") %>%
  group_by(gameId, playId) %>%
  summarise(
    actual_pressure = as.integer(any(pff_hit == 1 | pff_hurry == 1 | pff_sack == 1, na.rm = TRUE)),
    n_rushers       = n(),
    .groups = "drop"
  )

```

---

## 4. **PRSS(Pocket Rate of Space Shrinkage)**

```{r define-function}
relevant_positions <- c("QB", "OL", "DL", "LB", "TE", "CB", "S")

# k: rolling window size
compute_qb_area <- function(play_data, k) {

  # Find the snap frame and the pass/ sack / spike frame
  snap_vals    <- play_data$frameId[play_data$event == "ball_snap"]
  release_vals <- play_data$frameId[play_data$event %in%
                    c("pass_forward", "qb_sack", "qb_spike")]

  if (length(snap_vals) == 0 || length(release_vals) == 0) {
    return(tibble(qb = NA_character_, max_neg_change = NA_real_))
  }

  snap_frame    <- min(snap_vals,    na.rm = TRUE)
  release_frame <- min(release_vals, na.rm = TRUE)

  if (!is.finite(snap_frame) || !is.finite(release_frame)) {
    return(tibble(qb = NA_character_, max_neg_change = NA_real_))
  }

  play_data <- play_data %>%
    filter(frameId >= snap_frame, frameId <= release_frame)

  # QB name
  qb_name <- play_data %>%
    filter(officialPosition == "QB") %>%
    distinct(displayName) %>%
    pull() %>%
    paste(collapse = "/")

  frames <- unique(play_data$frameId)

  # 需要至少 k+1 帧才能计算滚动均值
  if (length(frames) < (k + 1)) {
    return(tibble(qb = qb_name, max_neg_change = NA_real_))
  }

  # 逐帧计算QB的Voronoi面积
  qb_areas <- map_dfr(frames, function(f) {
    frame_data <- play_data %>%
      filter(
        frameId == f,
        !is.na(x), !is.na(y),
        officialPosition %in% relevant_positions
      ) %>%
      distinct(x, y, .keep_all = TRUE)

    if (nrow(frame_data) < 3) return(tibble(frameId = f, qb_area = NA_real_))

    tryCatch({
      vor    <- deldir(frame_data$x, frame_data$y)
      vor_df <- vor$summary %>%
        select(x, y, dir.area) %>%
        left_join(frame_data, by = c("x", "y"))
      qb_area <- vor_df %>%
        filter(officialPosition == "QB") %>%
        pull(dir.area)
      tibble(frameId = f, qb_area = ifelse(length(qb_area) > 0, qb_area[1], NA_real_))
    }, error = function(e) tibble(frameId = f, qb_area = NA_real_))
  }) %>%
    arrange(frameId) %>%
    mutate(
      rate_change = c(NA, diff(qb_area)),
      rolling_avg = suppressWarnings(
        zoo::rollmean(rate_change, k = k, align = "left", fill = NA)
      )
    )

  max_neg <- suppressWarnings(min(qb_areas$rolling_avg, na.rm = TRUE))
  if (!is.finite(max_neg)) max_neg <- NA_real_

  tibble(qb = qb_name, max_neg_change = max_neg)
}
```

---

## 5. 对 k = 3, 5, 7, 10 批量计算

### k=5
```{r}
play_list <- tracking %>%
  group_by(gameId, playId) %>%
  nest()

results <- vector("list", nrow(play_list))

for (i in seq_len(nrow(play_list))) {
  if (i %% 100 == 0) message("Progress: ", i, "/", nrow(play_list))
  results[[i]] <- compute_qb_area(play_list$data[[i]], k = 5)
}

all_summary_k5 <- bind_rows(results) %>%
  bind_cols(play_list %>% select(gameId, playId))


# 跑完立刻保存，防止丢失
saveRDS(all_summary_k5, "../data/processed/all_summary_k5.rds")
```
```{r}
final_data <- all_summary_k5 %>%
  filter(!is.na(max_neg_change), !is.na(qb)) %>%
  left_join(
    plays %>% select(gameId, playId, possessionTeam, 
                     passResult, prePenaltyPlayResult),
    by = c("gameId", "playId")
  ) %>%
  left_join(pff_pressure, by = c("gameId", "playId"))
```

```{r}
# 画第一张图：PRSS 分布
ggplot(final_data, aes(x = max_neg_change)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of PRSS (k=5)",
    x = "PRSS (negative = pocket collapsing)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# 画第二张图：PRSS vs playResult
ggplot(final_data, aes(x = max_neg_change, y = prePenaltyPlayResult)) +
  geom_point(alpha = 0.2, color = "lightcoral") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "PRSS vs Play Result (yards gained)",
    x = "PRSS",
    y = "Yards Gained"
  ) +
  theme_minimal()

# 简单相关性
cor(final_data$max_neg_change, final_data$prePenaltyPlayResult, use = "complete.obs")
```

```{r}
# k=5 已有，直接复用
k5 <- all_summary_k5 %>% mutate(window = as.factor(5))

# 只跑 k=3, 7, 10
window_sizes <- c(3, 7, 10)
all_results <- vector("list", length(window_sizes))

for (w in seq_along(window_sizes)) {
  k <- window_sizes[w]
  message("Computing k = ", k, " ...")
  
  results_k <- vector("list", nrow(play_list))
  for (i in seq_len(nrow(play_list))) {
    if (i %% 500 == 0) message("  k=", k, " Progress: ", i, "/", nrow(play_list))
    results_k[[i]] <- compute_qb_area(play_list$data[[i]], k = k)
  }
  
  all_results[[w]] <- play_list %>%
    select(gameId, playId) %>%
    bind_cols(bind_rows(results_k)) %>%
    mutate(window = as.factor(k))
}

# 合并所有窗口包括k=5
all_summaries <- bind_rows(all_results) %>%
  bind_rows(k5)

saveRDS(all_summaries, "../data/processed/all_summaries_all_windows.rds")
```

## 6. review the k 
```{r}
# 读入
all_summaries <- readRDS("../data/processed/all_summaries_all_windows.rds")

# 合并 plays 数据
all_summaries_final <- all_summaries %>%
  filter(!is.na(max_neg_change), !is.na(qb)) %>%
  left_join(
    plays %>% select(gameId, playId, possessionTeam, 
                     passResult, prePenaltyPlayResult),
    by = c("gameId", "playId")
  )

# 看基本情况
all_summaries_final %>%
  group_by(window) %>%
  summarise(
    n = n(),
    na_count = sum(is.na(max_neg_change)),
    mean_prss = mean(max_neg_change, na.rm = TRUE),
    sd_prss = sd(max_neg_change, na.rm = TRUE)
  )
```

## PRSS visualization
```{r}
# 图1：四个窗口PRSS分布对比
ggplot(all_summaries_final, aes(x = max_neg_change, fill = window, color = window)) +
  geom_density(alpha = 0.25) +
  xlim(-20, 10) +  # 去掉极端值方便看
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "PRSS Distribution by Rolling Window Size",
    x = "PRSS",
    y = "Density",
    fill = "Window (k)",
    color = "Window (k)"
  ) +
  theme_minimal(base_size = 13)
```
As the rolling window size increases, the PRSS distribution shifts rightward and becomes wider.
k=3 (red): The distribution is very narrow and tall, centered just below 0. The small window picks up every small fluctuation in pocket area, but those fluctuations cancel each other out quickly, so the final PRSS values cluster tightly around 0.
k=5 (purple) and k=7 (green): Progressively wider and slightly more shifted to the right. These windows smooth out frame-to-frame noise while still capturing genuine pocket collapse events.
k=10 (teal): The widest and most right-shifted distribution, with mean PRSS actually turning positive (+0.46). The window is so large that it averages out real pressure events, making plays look like the pocket was stable or even expanding when it actually collapsed.

There is a tradeoff between sensitivity and smoothness. k=3 is too reactive to noise, k=10 over-smooths and loses real pressure signals. k=5 or k=7 appears to be the sweet spot — capturing genuine pocket collapse while filtering out frame-to-frame measurement noise.
The next step is to check which k value produces the strongest correlation with yards gained, which will help confirm the optimal window size.

```{r}

# 图2：各窗口与yards的相关性
all_summaries_final %>%
  group_by(window) %>%
  summarise(
    correlation = cor(max_neg_change, prePenaltyPlayResult, use = "complete.obs"),
    .groups = "drop"
  )

```
All four correlations are extremely weak (0.018 - 0.028), and there's no meaningful difference between window sizes.
What this tells us:
The choice of k barely matters for predicting yards gained at the play level. All windows perform essentially the same.

```{r}

# 只用k=5，filter掉scramble
prss_by_result <- all_summaries_final %>%
  filter(window == '5', passResult %in% c("C", "I", "S", "IN")) %>%
  mutate(passResult = factor(passResult, 
                             levels = c("S", "IN", "I", "C"),
                             labels = c("Sack", "Interception", "Incomplete", "Complete")))

# 箱线图
ggplot(prss_by_result, aes(x = passResult, y = max_neg_change, fill = passResult)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "PRSS Distribution by Pass Result (k=5)",
    x = "Pass Result",
    y = "PRSS (more negative = more pressure)",
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# 各组均值
prss_by_result %>%
  group_by(passResult) %>%
  summarise(
    mean_prss = mean(max_neg_change, na.rm = TRUE),
    median_prss = median(max_neg_change, na.rm = TRUE),
    n = n()
  )

```
## 6. 合并外部数据

```{r merge-data, message=FALSE}
library(nflfastR)

pbp <- load_pbp(2021) %>%
  mutate(
    gameId = as.numeric(old_game_id),
    playId = as.numeric(play_id)
  ) %>%
  select(gameId, playId, epa, yards_gained, cpoe, air_yards)


all_summaries_final <- all_summaries_final %>%
  left_join(pbp, by = c("gameId", "playId"))

# 确认join成功
sum(!is.na(all_summaries_final$epa))
```

## 7. prss vs epa
```{r}

all_summaries_final %>%
  group_by(window) %>%
  summarise(
    correlation = cor(max_neg_change, epa, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )
```
At the individual play level, PRSS shows very weak correlation with EPA across all rolling window sizes (r = 0.04–0.05), with no meaningful difference between k = 3, 5, 7, and 10.

This is not necessarily an indicator that PRSS is ineffective. Pocket geometry is only one of many factors that determine the outcome of a play — receiver route quality, defensive coverage, quarterback decision-making, and receiver performance all contribute independently to EPA. A quarterback may face severe pocket collapse yet still deliver a successful pass, or face minimal pressure yet throw an incompletion.

The consistency of correlations across all four window sizes also suggests that the choice of k does not meaningfully affect PRSS at the play level, implying the metric is relatively robust to this parameter.

The more appropriate level of analysis may be at the QB or team level, where averaging across many plays reduces the noise from individual play variance and allows the signal from pocket pressure to emerge more clearly.
---


